### 2017-12-04 paper/sexSelectionSpacing-ver1.tex
Edit abstract and introduction
Abstract:
- Emphasize changes in spacing over time

Introduction:
- General edits
- Highlight why spacing is still important to understand

\date{October 2017\\
\bigskip
Preliminary}

@@

% Demography abstract
\noindent This paper examines how birth spacing interact with the use of 
sex-selective abortions. 
I introduce a statistical method that incorporates that sex-selective abortions 
affect both the likelihood of a son and spacing between births.
Using India's National Family and Health Surveys,
I show that use of sex selection leads to \emph{longer} spacing after a daughter 
than after a son, but for families that do not use sex selection
duration to next birth is still shorter when no boys are present.
Women with 8 or more years of education, both in urban and rural areas, are 
the main users of sex-selective abortions, whereas women with less education 
do not appear to use sex selection.

@@

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
This is because each sex-selective abortion substantially increases the 
spacing from last birth. 
For every abortion a woman has we should expect the duration from last
birth to increase by approximately a year.
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter
because of the use of sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference. 
Spacing is, however, still important and useful in understanding son preference, 
but only if combined with the likelihood of observing a boy or a girl.
Short spacing after a girl is indicative of son preference if observed
births are close to the natural sex ratio, whereas long spacing is indicative
of son preference if observed births are male biased.

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, even 
parents with strong son preference may reverse their decision to use 
prenatal sex determination, and carry the next pregnancy to term whether 
male or female.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is, but the 
first reports of on the availability of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

>>>>

\date{December 2017\\
\bigskip
Preliminary}

@@

% Demography abstract
\noindent

Strong son preference has frequently been associated with shorter 
spacing between births after the birth of a girl.
The spread of access to prenatal sex determination and 
sex-selective abortions has the potential to reverse this pattern
because each abortion adds substantially to the duration between births.
I introduce a statistical method that simultaneously account for how sex 
selection increases the likelihood of a son \emph{and} the spacing between births.
Using India's National Family and Health Surveys,
I show that use of sex selection leads to \emph{longer} spacing after a daughter 
than after a son; a reversal of the pattern before the availability of sex selection.
For families that do not use sex selection, the duration to next birth is still 
shorter when no boys are present.

Women with 8 or more years of education, both in urban and rural areas, are 
the main users of sex-selective abortions, whereas women with less education 
do not appear to use sex selection.

@@

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
Couples with stronger son preferences are more likely to use prenatal
sex determination and sex-selective abortions, but each sex-selective 
abortion substantially increases the duration from last birth. 
Each abortion increases spacing between births by approximately a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of a spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which I set to 6 
months \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception varies by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter,
because of their use of sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference.
Understanding birth spacing remains, however, an important undertaking.
First, spacing is still important and useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
% Short spacing after a girl is indicative of son preference if observed
% births are close to the natural sex ratio, whereas long spacing is indicative
% of son preference if observed births are male biased.
Second, if spacing affect health outcomes for mother and/or children, it is 
important to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arise
from sex-selective abortions rather than be driven by a higher value of
girls resulting from lower supply \citep{Lin2014,Hu2015}.
Third, spacing may be an important factor in parents' decisions, either
for preference or economic reasons, such as labor force participation.
This is especially important here because even parents with strong
son preference may reverse their decision to use prenatal sex determination, 
and carry the next pregnancy to term whether male or female, as the duration 
from last birth becomes sufficiently sufficiently long.%
\footnote{
More generally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be important
for the design of suitable policies \citep{Portner2018}.
}

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971, and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.
